```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

`bionemor` provides R interfaces to selected biological foundation-model assets
and runtimes published through NVIDIA BioNeMo Recipes. BioNeMo Recipes supplies
family-specific programs and the runtime environments needed to run them. Each
model family has its own adapter, checkpoints, and runtime. The package supplies
the shared R-facing workflow: sequence inputs, compute configuration, durable
jobs, provenance, and R results.

> Model operations require a supported CUDA-capable NVIDIA GPU. There is no CPU fallback.
> You can install the package and inspect recipes, workflows, and model metadata
> without a GPU.

## How the pieces fit together

The package currently supports Evo 2 for DNA and ESM-2 for proteins. These are
separate model families, not variants of one model. They share the surrounding
work needed to run a biological model on a GPU, which is why they live in one
package.

| Concept | Role in bionemor |
|---|---|
| Model descriptor | A lightweight S7 object that identifies a model family, size, configuration, and optional checkpoint. It does not load model weights into the R process. |
| Checkpoint | Model weights stored outside R. A model descriptor can point to a checkpoint prepared or fitted in an earlier run. |
| Recipe | A versioned description of the family-specific source, dependencies, and commands used by an adapter. |
| Compute descriptor | The recipe plus the workspace, execution backend, runtime engine, and requested GPU resources. Models retain this descriptor when it is supplied. |
| Container | One way a compute descriptor supplies the recipe runtime. Local container work uses Docker by default; Slurm container work uses Apptainer. An externally managed runtime is the other engine. |
| Workflow | One family-specific operation, such as Evo 2 scoring or ESM-2 embedding. |
| Job | The durable record of one workflow execution, including its request, logs, state, outputs, and provenance. |

A model, sequence input, and operation define the requested work. The compute
descriptor says where and how to run it; its recipe supplies the matching
family-specific runtime. The operation produces a durable job and, when it
finishes, an ordinary R result.

## Find a workflow

Start with `bionemo_workflows()`. It lists the model families and operations in
the installed package without downloading model weights:

```{r workflows}
library(bionemor)

workflows <- bionemo_workflows()
workflows[c("id", "family", "task")]
```

The package currently provides these family-specific R interfaces:

| Model family | Input | Supported workflows | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Checkpoint preparation and export, generation, scoring, positional profiles, embeddings, training-data preprocessing, and fine-tuning | `evo2_models()`, `evo2_model()`, `evo2_checkpoint()`, `evo2_export()`, `evo2_prepare()`, `evo2_generate()`, `evo2_score()`, `evo2_profile()`, `evo2_embed()`, `evo2_finetune()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

Evo 2 has the broadest interface and is the detailed example below. ESM-2
embedding uses the same compute, workflow, and durable-job abstractions. The
lower-level `bionemo_run()` interface also accepts an ID from
`bionemo_workflows()` when you need to select a workflow programmatically.

The captured output below was rendered on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

Install the package on the Linux GPU machine where you will run the model.

## Get access to a GPU

The package-managed local runtime requires Linux, Git, `tar`, Docker, and the
NVIDIA Container Toolkit. Docker must be able to expose the GPU to containers.
The Evo 2 7B examples below were captured on one NVIDIA L40S with 48 GB of GPU
memory. Requirements vary by model, sequence length, batch size, and operation.

If you do not have a suitable local GPU, [NVIDIA Brev](https://brev.nvidia.com/)
is one way to rent one. Brev instances are billable. Inspect the available type
and hourly price before creating an instance:

```bash
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --stoppable --gpu-name L40S --min-vram 48
brev shell bionemor-gpu
```

Brev supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.
Inside the instance, use the `local` backend shown below. Install R and the
required system tools if the selected image does not include them.

Leave the remote shell and stop the instance from your local terminal when you
finish:

```bash
brev stop bionemor-gpu
```

Check the selected instance type's billing and storage terms.

## Install a recipe runtime

NVIDIA NGC is the container registry that supplies the runtime's base image.
Create an NGC API key and authenticate Docker to `nvcr.io` on the GPU machine,
including when using Brev:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

A compute descriptor records where jobs run, where their files are stored, and
which recipe supplies the model-specific runtime. The recipe is required so
that changing model families is visible in user code. This example uses a local
Docker container. `backend = "slurm"` also accepts an administrator-managed
runtime or an existing Apptainer image.

This descriptor selects the Evo 2 recipe:

```{r runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/bionemor-work"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

`bionemo_install()` builds or verifies the selected runtime and performs
GPU-backed capability checks. Use `bionemo_doctor()` to diagnose a setup
problem. A site-managed recipe environment can use `engine = "external"`.

## Evo 2: score, generate, and embed DNA

`evo2_models()` lists models without downloading weights. With a configured
compute object, it can filter by GPU count, compute capability, and precision
policy. This compatibility check does not estimate GPU memory or disk use.

```{r models}
models <- evo2_models(evo2_compute, compatible = TRUE)
str(models)
```

Megatron Bridge is the training and inference stack used by the pinned Evo 2
recipe. MBridge is its checkpoint format. `evo2_model()` needs a compute
descriptor because it downloads the registered source weights when needed,
converts them in the selected runtime, and stores the result in the compute
workspace. It returns a model descriptor that points to the prepared checkpoint
and reuses a complete matching checkpoint on later calls:

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r model-capture, include=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
```

The returned model retains its compute object. Generation and scoring return
data frames; pooled embeddings return a numeric matrix:

```{r evo2-inference}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores

dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)
dim(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Scores are reduced token log probabilities; higher values indicate greater
model likelihood. They are not calibrated measurements of biological
function. Embedding row names preserve the input IDs.

## ESM-2: embed proteins

ESM-2 uses its own recipe and compute descriptor. `esm2_model()` binds a pinned
ESM-2 checkpoint to that compute target. On first use, Transformers downloads
the selected weights from Hugging Face and caches them below the workspace.
Installation uses the recipe's native Transformers and Transformer Engine path
and does not compile vLLM. ESM-2 inference currently requires `gpus = 1`. You
may supply multiple proteins; the runtime processes them one at a time on the
selected GPU to preserve their bidirectional attention boundaries.

```{r esm2}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  gpus = 1L,
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is one last-token, L2-normalized protein embedding. The result remains
a standard numeric matrix for use with ordinary R tools; its `provenance`
attribute records the model, recipe revision, and durable run path. You can use
these vectors for sequence similarity, clustering, or as features in downstream
R models. They are model representations, not measurements of protein function.
`esm2_models()` lists the available pinned checkpoints and their embedding
dimensions without downloading weights.

## Run longer work

Family-specific functions use the same durable job interface. Set
`async = TRUE`, retain the job path, and reopen it in a later R session:

```r
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  async = TRUE
)

path <- job_path(run)

# In a new R session:
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

## More guides

- [BioNeMo workflows from R](https://t-kalinowski.github.io/bionemor/articles/bionemor.html)
  explains the shared API, then walks through Evo 2 and ESM-2 workflows.
- [Fine-tune Evo 2 and retain the checkpoint](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
  covers preprocessing, LoRA, and fitted inference.

Generated sequences are model output, not validated biological designs. The
package reports mechanical sequence checks; downstream biological validation
remains the user's responsibility.
