```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

`bionemor` prepares Evo 2 models, runs batch inference, and fine-tunes with
NVIDIA BioNeMo Recipes without leaving R. Inputs and results are R vectors,
data frames, matrices, and durable job objects.

The main workflow is:

1. Configure and verify a pinned recipe runtime once.
2. Prepare or reuse the recommended checkpoint with `evo2_model()`.
3. Generate, score, embed, or fine-tune DNA sequences.

The output below was rendered on `r docs_date` with an `r docs_gpu` using
package revision `r substr(docs_revision, 1L, 12L)`. See
[`vignettes-src/`](vignettes-src/) and
[`tools/render-gpu-docs.R`](tools/render-gpu-docs.R) for the executable sources
and manual render command.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

## Set up the recipe runtime once

The default local setup builds the verified Evo 2 recipe container. It requires
Git, `tar`, Docker, and an NVIDIA GPU exposed through Docker.

Authenticate Docker to `nvcr.io` with an NGC API key before installing:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

Then inspect the setup plan, install the runtime, and verify it:

```{r runtime}
library(bionemor)

workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/evo2-work"),
  mustWork = FALSE
)
compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)

plan <- bionemo_install_plan(compute)
plan

compute <- bionemo_install(compute)
doctor <- bionemo_doctor(compute, target = "all", verbose = FALSE)
stopifnot(doctor@ok)
doctor
```

`bionemo_install()` performs GPU-backed capability and command probes. A
site-managed recipe environment can use `engine = "external"`.

## Prepare a model

`evo2_models()` lists the registry and compatibility metadata without preparing
weights:

```{r models}
models <- evo2_models(compute, compatible = TRUE)
models[models$name %in% c("1b-base", "7b"), c(
  "name", "parameters", "context_length", "compatible", "download_size"
)]
```

`evo2_model()` prepares the recommended dense MBridge checkpoint on its first
call and reuses a complete matching checkpoint later:

```r
model <- evo2_model("7b", compute)
```

An already prepared custom checkpoint can be attached and compute-bound without
copying or converting it. The rendered capture uses this path:

```{r model}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = compute
)
model
```

The model now retains its compute descriptor. Inference and fine-tuning calls
therefore do not need a repeated `compute` argument. The separation still
matters at construction: a model describes architecture and weights, while
`bionemo_compute()` describes the runtime, workspace, backend, recipe revision,
and GPU allocation. `evo2_model()` couples the recommended checkpoint and
compute; `evo2(..., compute = compute)` does the same for a custom checkpoint.

## Run inference

Generation and scoring return data frames. Pooled embeddings return a numeric
matrix:

```{r inference}
sequences <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  sequences,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]

scores <- evo2_score(
  model,
  sequences,
  reduction = "mean",
  strand = "both"
)
scores[c("id", "score", "forward_score", "reverse_score")]

embeddings <- evo2_embed(
  model,
  sequences,
  pool = "mean",
  strand = "both"
)
dim(embeddings)
round(embeddings[, 1:4, drop = FALSE], 4)
```

Scores are reduced token log probabilities; higher values indicate greater
model likelihood. With `strand = "both"`, forward and reverse values are
reduced independently and averaged. Embedding row names preserve the input IDs.

## Fine-tune with LoRA

Raw R sequence vectors are prepared automatically. This small run exists to
exercise the API; two optimizer steps on repeated synthetic sequences do not
produce a useful biological model.

```{r fine-tune}
sequence_127 <- function(pattern) {
  substr(strrep(pattern, ceiling(127 / nchar(pattern))), 1L, 127L)
}

data <- evo2_dataset(
  train = c(
    train_1 = sequence_127("ACGT"),
    train_2 = sequence_127("TGCA"),
    train_3 = sequence_127("GATTACA"),
    train_4 = sequence_127("CCGTA")
  ),
  validation = c(validation_1 = sequence_127("AGCT")),
  test = c(test_1 = sequence_127("TGCAT"))
)
run_id <- Sys.getenv(
  "BIONEMOR_DOCS_RUN_ID",
  format(Sys.time(), "docs-%Y%m%d-%H%M%S")
)

run <- evo2_finetune(
  model,
  data,
  steps = 2L,
  method = evo2_lora(
    rank = 4L,
    alpha = 8,
    dropout = 0,
    targets = c("hyena", "attention", "mlp")
  ),
  control = evo2_fit_control(
    sequence_length = 128L,
    global_batch_size = 1L,
    micro_batch_size = 1L,
    learning_rate = 1e-4,
    minimum_learning_rate = 0,
    warmup_steps = 0L,
    decay_steps = 2L,
    constant_steps = 0L,
    eval_interval = 1L,
    eval_iters = 1L,
    log_interval = 1L,
    precision = "bf16",
    keep_checkpoints = 1L,
    workers = 1L,
    seed = 17L,
    dataset_seed = 17L
  ),
  path = file.path("artifacts", run_id, "readme-lora"),
  name = paste0(run_id, "-readme-lora"),
  async = TRUE,
  timeout = 3600
)

fitted <- job_wait(run, timeout = 3600)
job_status(run, refresh = TRUE)
fitted

manifest <- checkpoint_manifest(fitted)
data.frame(
  kind = manifest$kind,
  format = manifest$format,
  base_checkpoint = basename(manifest$base_checkpoint_path),
  recipe_revision = substr(manifest$recipe_revision, 1L, 12L)
)
```

`targets` selects module groups in every matching model layer. `"hyena"`
expands to `dense_projection` and `dense`; `"attention"` expands to
`linear_qkv` and `linear_proj`; and `"mlp"` expands to `linear_fc1` and
`linear_fc2`. The effective LoRA scale is `alpha / rank`.

The fitted result is another compute-bound `Evo2Model`:

```{r fitted-inference}
probe <- c(probe = sequence_127("ACGTGCAA"))

evo2_score(
  fitted,
  probe,
  reduction = "mean",
  strand = "forward"
)[c("id", "score")]

evo2_generate(
  fitted,
  probe,
  num_tokens = 8L,
  seed = 17L
)[c("input_id", "prompt", "completion", "finish_reason")]
```

The LoRA checkpoint records its dense base checkpoint. Keep both paths
available for later inference.

## Vignettes and durable runs

The package includes three articles:

- [Evo 2 workflows from R](vignettes/bionemor.Rmd)
- [Fine-tune Evo 2 and retain the checkpoint](vignettes/evo2-finetune.Rmd)
- [Run BioNeMo Recipes jobs with Slurm](vignettes/slurm.Rmd)

The first two contain output captured in the target GPU environment. The Slurm
article remains a reference because scheduler behavior must be validated on the
target cluster. Slurm support is experimental.

Every operation stores its request, command plan, state, logs, outputs, and
provenance under the compute workspace. Reopen a run after the launching R
session exits:

```r
same_run <- bionemo_job(job_path(run))
job_status(same_run)
result <- job_wait(same_run)
```

Generated sequence is model output, not a validated biological design. The
package reports mechanical sequence checks; downstream biological validation
remains the user's responsibility.
