```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

Biological foundation models learn from large collections of DNA or protein
sequences. The resulting models can be used across different sequence tasks.

`bionemor` lets bioinformaticians and other researchers use these models on
NVIDIA GPUs from R. With Evo 2, you can generate and score DNA, calculate
positional profiles and embeddings, and fine-tune a model on your own sequences.
With ESM-2, you can create protein embeddings for similarity, clustering, and
downstream models. Embeddings are numerical representations of sequences.

Sequence inputs can be character vectors or sequence sets from Biostrings.
Generation and scoring return data frames, while pooled embeddings return
numeric matrices. Longer runs can be saved, monitored, and reopened from another
R session.

> Running a model requires a supported CUDA-capable NVIDIA GPU. There is no CPU fallback.
> You can install `bionemor` and inspect the available models without a GPU.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

For the local container setup below, install the package in the R environment on
the Linux GPU machine, then load it:

```{r package}
library(bionemor)
```

## Supported models and operations

The supported models and their main functions are:

| Model family | Input | What you can do | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Generation, scoring, positional profiles, embeddings, fine-tuning, training-data preprocessing, and checkpoint preparation and export | `evo2_generate()`, `evo2_score()`, `evo2_profile()`, `evo2_embed()`, `evo2_finetune()`, `evo2_preprocess()`, `evo2_models()`, `evo2_model()`, `evo2_checkpoint()`, `evo2_export()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

The examples below begin with Evo 2 DNA, followed by ESM-2 protein embeddings.

## Get access to a GPU

For local container execution, run R on a Linux machine with Git, `tar`, Docker,
and the NVIDIA Container Toolkit. Docker must be configured to expose the GPU
to containers. GPU requirements vary by model, sequence length, batch size, and
operation.

If you do not have a suitable local GPU, [NVIDIA Brev](https://brev.nvidia.com/)
is one way to rent one. Brev instances are billable; review the current hourly
price and storage terms before creating an instance:

```bash
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --stoppable --gpu-name L40S --min-vram 48
brev shell bionemor-gpu
```

Inside the Brev instance, install R and the required system tools if the
selected image does not include them, then follow the local setup below.

Leave the remote shell and stop the instance from your local terminal when you
finish:

```bash
brev stop bionemor-gpu
```

## Set up Evo 2

Evo 2 operations run through NVIDIA BioNeMo Recipes and Megatron Bridge.
`bionemor` uses versions tested with this release. The local Docker environment
is built from an NVIDIA NGC base image. Create an NGC API key and authenticate
Docker to `nvcr.io` on the GPU machine, including when using Brev:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

`bionemo_compute()` records the workspace and execution settings.
`evo2_recipe()` selects that software environment; `backend = "local"` and
`engine = "container"` run it in Docker on the current machine:

```{r runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/bionemor-work"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

`bionemo_install()` builds or verifies the selected environment and checks its
GPU capabilities. If setup fails, use `bionemo_doctor()` to inspect the
environment.

## Evo 2: generate, score, and embed DNA

`evo2_models()` lists models without downloading weights. With a configured
compute object, it can filter by GPU count, compute capability, and precision
policy. Check GPU memory and disk requirements separately.

The output below was captured on `r docs_date` with an `r docs_gpu` using
package revision `r substr(docs_revision, 1L, 12L)`.

```{r models}
models <- evo2_models(evo2_compute, compatible = TRUE)
str(models)
```

`evo2_model()` downloads and prepares the selected model the first time you use
it. Later calls from the same workspace reuse the prepared checkpoint:

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r model-capture, include=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
```

Pass the model to the generation, scoring, and embedding functions. Generation
and scoring return data frames; pooled embeddings return a numeric matrix:

```{r evo2-inference}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores

dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)
dim(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Scores are reduced token log probabilities, not calibrated measurements of
biological function. Higher values indicate greater model likelihood. Generated
sequences are model outputs and require downstream biological validation.
Embedding row names preserve the input IDs.

## ESM-2: embed proteins

`bionemor` supports pooled protein embeddings with ESM-2. You can use these
vectors for sequence similarity, clustering, or as features in downstream R
models. `esm2_models()` lists the available pinned NVIDIA checkpoints and their
embedding dimensions without downloading weights, and `esm2_model()` selects
one for use with a compute descriptor.

The ESM-2 environment uses Transformers. On first use, Transformers downloads
the selected weights from Hugging Face and caches them below the workspace.
`esm2_embed()` currently requires `gpus = 1` and returns one embedding per
protein:

```{r esm2}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  gpus = 1L,
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is one last-token, L2-normalized protein embedding. The result is a
numeric matrix for use with ordinary R tools. Its `provenance` attribute records
the model, recipe revision, and path where the job was saved. These embeddings
are model representations, not measurements of protein function.

## Run longer work

Operations that accept `async = TRUE` return a saved job. Save its path to
monitor it or reopen it in a later R session:

```r
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  async = TRUE
)

path <- job_path(run)

# In a new R session:
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

## How the pieces fit together

The examples above use the following package concepts:

| Concept | Role in bionemor |
|---|---|
| Model descriptor | Identifies the model family, size, configuration, and optional checkpoint. |
| Checkpoint | Model weights obtained from a registered source, a supplied path, or an earlier run. |
| Recipe | Describes the versioned software environment and commands for a model family. |
| Compute descriptor | Combines a recipe with a workspace, execution settings, and requested GPU resources. |
| Operation | Work requested through a family-specific R function such as `evo2_score()` or `esm2_embed()`. |
| Job | A saved record of an operation's request, logs, state, outputs, and provenance. |

## More guides

- [Get started with bionemor](https://t-kalinowski.github.io/bionemor/articles/bionemor.html)
  walks through setup and Evo 2 and ESM-2 examples.
- [Fine-tune Evo 2 and retain the checkpoint](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
  covers preprocessing, LoRA, and fitted inference.
