```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

`bionemor` runs biological foundation-model workflows from R with NVIDIA
BioNeMo Recipes. BioNeMo Recipes is a collection of model-specific training and
inference programs. The package installs a pinned recipe runtime, translates R
inputs into its command-line interface, and returns results as R objects and
durable jobs.

> Model operations require a supported CUDA-capable NVIDIA GPU. There is no CPU fallback.
> You can install the package and inspect recipes, workflows, and model metadata
> without a GPU.

## Find a workflow

Start with `bionemo_workflows()`. It lists the model families and operations in
the installed package without downloading model weights:

```{r workflows}
library(bionemor)

workflows <- bionemo_workflows()
workflows[c("id", "family", "task")]
```

The package currently provides these family-specific R interfaces:

| Model family | Input | Supported workflows | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Checkpoint preparation and export, generation, scoring, positional profiles, embeddings, data preparation, and fine-tuning | `evo2_models()`, `evo2_model()`, `evo2_generate()`, `evo2_score()`, `evo2_embed()`, `evo2_finetune()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

Evo 2 has the broadest interface and is the detailed example below. ESM-2
embedding uses the same compute, workflow, and durable-job abstractions. The
lower-level `bionemo_run()` interface also accepts an ID from
`bionemo_workflows()` when you need to select a workflow programmatically.

The captured output below was rendered on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

Install the package on the Linux GPU machine where you will run the model.

## Get access to a GPU

The package-managed local runtime requires Linux, Git, `tar`, Docker, and the
NVIDIA Container Toolkit. Docker must be able to expose the GPU to containers.
The Evo 2 7B examples below were captured on one NVIDIA L40S with 48 GB of GPU
memory. Requirements vary by model, sequence length, batch size, and operation.

If you do not have a suitable local GPU, [NVIDIA Brev](https://brev.nvidia.com/)
is one way to rent one. Brev instances are billable. Inspect the available type
and hourly price before creating an instance:

```bash
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --stoppable --gpu-name L40S --min-vram 48
brev shell bionemor-gpu
```

Brev supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.
Inside the instance, use the `local` backend shown below. Install R and the
required system tools if the selected image does not include them.

Leave the remote shell and stop the instance from your local terminal when you
finish:

```bash
brev stop bionemor-gpu
```

Check the selected instance type's billing and storage terms.

## Install a recipe runtime

NVIDIA NGC is the container registry that supplies the runtime's base image.
Create an NGC API key and authenticate Docker to `nvcr.io` on the GPU machine,
including when using Brev:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

A compute descriptor records where jobs run, where their files are stored, and
which recipe supplies the model-specific runtime. The recipe is required so
that changing model families is visible in user code.

This descriptor selects the Evo 2 recipe:

```{r runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/bionemor-work"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

`bionemo_install()` builds or verifies the selected runtime and performs
GPU-backed capability checks. Use `bionemo_doctor()` to diagnose a setup
problem. A site-managed recipe environment can use `engine = "external"`.

## Evo 2: score, generate, and embed DNA

`evo2_models()` lists models without downloading weights. With a configured
compute object, it can filter by GPU count, compute capability, and precision
policy. This compatibility check does not estimate GPU memory or disk use.

```{r models}
models <- evo2_models(evo2_compute, compatible = TRUE)
str(models)
```

`evo2_model()` prepares the recommended dense Megatron Bridge checkpoint on
its first call and reuses a complete matching checkpoint later:

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r model-capture, include=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
```

The returned model retains its compute object. Generation and scoring return
data frames; pooled embeddings return a numeric matrix:

```{r evo2-inference}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores

dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)
dim(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Scores are reduced token log probabilities; higher values indicate greater
model likelihood. They are not calibrated measurements of biological
function. Embedding row names preserve the input IDs.

## ESM-2: embed proteins

ESM-2 uses its own recipe and compute descriptor. `esm2_model()` binds a pinned
ESM-2 checkpoint to that compute target. On first use, vLLM downloads the
selected weights from Hugging Face and caches them below the workspace.
During installation, the package-managed image compiles vLLM's CUDA/C++
extensions from source for the current NVIDIA GPU architecture. On an 8-vCPU
machine, this first build can take tens of minutes; later calls reuse the image.
Rebuild it before using a GPU with a different compute capability.

```{r esm2}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is one last-token, L2-normalized protein embedding. You can use these
vectors for sequence similarity, clustering, or as features in downstream R
models. They are model representations, not measurements of protein function.
`esm2_models()` lists the available pinned checkpoints and their embedding
dimensions without downloading weights.

## Run longer work

Family-specific functions use the same durable job interface. Set
`async = TRUE`, retain the job path, and reopen it in a later R session:

```r
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  async = TRUE
)

path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

## More guides

- [BioNeMo workflows from R](vignettes/bionemor.Rmd) explains the shared API,
  then walks through Evo 2 and ESM-2 workflows.
- [Fine-tune Evo 2 and retain the checkpoint](vignettes/evo2-finetune.Rmd)
  covers preprocessing, LoRA, and fitted inference.
- [Run BioNeMo Recipes jobs with Slurm](vignettes/slurm.Rmd) is an experimental
  cluster reference. Its Evo 2 example has not been executed on Slurm and must
  be validated on the target cluster.

Generated sequences are model output, not validated biological designs. The
package reports mechanical sequence checks; downstream biological validation
remains the user's responsibility.
