```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

Biological foundation models learn from large collections of DNA or protein
sequences, then can be reused across different sequence tasks. They can generate
or score sequences, create numerical representations called embeddings for
similarity, clustering, or downstream models, and be adapted to a particular
collection of sequences through fine-tuning.

`bionemor` lets bioinformaticians run these operations from R on NVIDIA GPUs. It
is intended for people whose sequence data and downstream analysis already live
in R, including users of Bioconductor packages such as Biostrings. The package
currently supports Evo 2 for DNA generation, scoring, positional profiles,
embeddings, preprocessing, and LoRA or full fine-tuning. It also supports ESM-2
for pooled protein embeddings.

For each operation it exposes, `bionemor` runs the pinned upstream model code and
checkpoints. The model computation is the same as in the upstream runtime, not
a simplified R implementation. Evo 2 uses NVIDIA BioNeMo Recipes and Megatron
Bridge. ESM-2 uses NVIDIA checkpoints with Transformers and Transformer Engine.

`bionemor` adds an R-native orchestration and results layer around those
runtimes. It configures compute, builds or checks runtimes, stages sequence
inputs, prepares and reuses checkpoints, runs and monitors jobs, returns
portable R results, and records how they were produced. This covers the path
from an R or Bioconductor sequence object to GPU computation and back to an R
object for analysis.

> Model operations require a supported CUDA-capable NVIDIA GPU. There is no CPU fallback.
> You can install the package and inspect recipes and model metadata
> without a GPU.

## Supported models and operations

```{r package}
library(bionemor)
```

The function name identifies the model family and the work being requested:

| Model family | Input | Available operations | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Checkpoint preparation and export, generation, scoring, positional profiles, embeddings, training-data preprocessing, and fine-tuning | `evo2_models()`, `evo2_model()`, `evo2_checkpoint()`, `evo2_export()`, `evo2_preprocess()`, `evo2_generate()`, `evo2_score()`, `evo2_profile()`, `evo2_embed()`, `evo2_finetune()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

Use these family-specific functions directly. The examples below begin with Evo
2 DNA, then show ESM-2 protein embeddings through the same R interface.

The captured output below was rendered on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

Install the package on the Linux GPU machine where you will run the model.

## Get access to a GPU

The package-managed local runtime requires Linux, Git, `tar`, Docker, and the
NVIDIA Container Toolkit. Docker must be able to expose the GPU to containers.
The Evo 2 7B examples below were captured on one NVIDIA L40S with 48 GB of GPU
memory. Requirements vary by model, sequence length, batch size, and operation.

If you do not have a suitable local GPU, [NVIDIA Brev](https://brev.nvidia.com/)
is one way to rent one. Brev instances are billable. Inspect the available type
and hourly price before creating an instance:

```bash
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --stoppable --gpu-name L40S --min-vram 48
brev shell bionemor-gpu
```

Brev supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.
Inside the instance, use the `local` backend shown below. Install R and the
required system tools if the selected image does not include them.

Leave the remote shell and stop the instance from your local terminal when you
finish:

```bash
brev stop bionemor-gpu
```

Check the selected instance type's billing and storage terms.

## Install a recipe runtime

NVIDIA NGC is the container registry that supplies the runtime's base image.
Create an NGC API key and authenticate Docker to `nvcr.io` on the GPU machine,
including when using Brev:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

A compute descriptor records where jobs run, where their files are stored, and
which recipe supplies the model-specific runtime. The recipe is required so
that changing model families is visible in user code. This example uses a local
Docker container. `backend = "slurm"` also accepts an administrator-managed
runtime or an existing Apptainer image.

This descriptor selects the pinned NVIDIA BioNeMo Recipes runtime for Evo 2:

```{r runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/bionemor-work"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

`bionemo_install()` builds or verifies the selected runtime and performs
GPU-backed capability checks. Use `bionemo_doctor()` to diagnose a setup
problem. A site-managed recipe environment can use `engine = "external"`.

## Evo 2: score, generate, and embed DNA

`evo2_models()` lists models without downloading weights. With a configured
compute object, it can filter by GPU count, compute capability, and precision
policy. This compatibility check does not estimate GPU memory or disk use.

```{r models}
models <- evo2_models(evo2_compute, compatible = TRUE)
str(models)
```

Megatron Bridge is the training and inference stack used by the pinned Evo 2
recipe. MBridge is its checkpoint format. `evo2_model()` needs a compute
descriptor because it downloads the registered source weights when needed,
converts them in the selected runtime, and stores the result in the compute
workspace. It returns a model descriptor that points to the prepared checkpoint
and reuses a complete matching checkpoint on later calls:

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r model-capture, include=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
```

The returned model retains its compute object. Generation and scoring return
data frames; pooled embeddings return a numeric matrix:

```{r evo2-inference}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores

dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)
dim(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Scores are reduced token log probabilities; higher values indicate greater
model likelihood. They are not calibrated measurements of biological
function. Embedding row names preserve the input IDs.

## ESM-2: embed proteins

`bionemor` also supports ESM-2, a protein language model for representing
protein sequences. `esm2_model()` selects a pinned ESM-2 checkpoint and binds it
to a compute descriptor. `esm2_embed()` returns one pooled embedding per
protein. On first use, Transformers downloads the selected weights from Hugging
Face and caches them below the workspace. The package installs the recipe's
native Transformers and Transformer Engine runtime. ESM-2 inference currently
requires `gpus = 1`. You may supply multiple proteins; the runtime processes
them one at a time on the selected GPU to preserve their bidirectional attention
boundaries.

```{r esm2}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  gpus = 1L,
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is one last-token, L2-normalized protein embedding. The result remains
a standard numeric matrix for use with ordinary R tools; its `provenance`
attribute records the model, recipe revision, and path where the job was saved.
You can use these vectors for sequence similarity, clustering, or as features in
downstream R models. They are model representations, not measurements of protein
function.
`esm2_models()` lists the available pinned checkpoints and their embedding
dimensions without downloading weights.

## How the pieces fit together

The Evo 2 and ESM-2 examples above use the same R building blocks. A model and
checkpoint identify what will run; a recipe and compute descriptor say where
and how to run it; and a job records the work and returns its result to R.

| Concept | Role in bionemor |
|---|---|
| Model descriptor | A lightweight S7 object that identifies a model family, size, configuration, and optional checkpoint. It does not load model weights into the R process. |
| Checkpoint | Model weights stored outside R. A model descriptor can point to a checkpoint prepared or fitted in an earlier run. |
| Recipe | A versioned description of the family-specific source, dependencies, and commands used to build or verify its runtime. |
| Compute descriptor | The recipe plus the workspace, execution backend, runtime engine, and requested GPU resources. Models retain this descriptor when it is supplied. |
| Container | One way a compute descriptor supplies the recipe runtime. Local container work uses Docker by default; Slurm container work uses Apptainer. An externally managed runtime is the other engine. |
| Operation | A family-specific R function, such as `evo2_score()` or `esm2_embed()`, that requests work from a model. |
| Job | A record of one operation, including its request, logs, state, outputs, and details about how it was run. It is saved on disk and can be reopened in another R session. |

A family-specific function combines a model with the relevant input. The
compute descriptor says where and how to run it; its recipe supplies the
matching runtime. Every operation is recorded in a job directory. Synchronous
calls return an ordinary R result; asynchronous calls return a job that can be
monitored or reopened in another R session.

## Run longer work

Family-specific functions use the same job interface. Set `async = TRUE`, save
the job path, and reopen it in a later R session:

```r
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  async = TRUE
)

path <- job_path(run)

# In a new R session:
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

## More guides

- [Get started with bionemor](https://t-kalinowski.github.io/bionemor/articles/bionemor.html)
  explains the shared API, then walks through Evo 2 and ESM-2 operations.
- [Fine-tune Evo 2 and retain the checkpoint](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
  covers preprocessing, LoRA, and fitted inference.

Generated sequences are model output, not validated biological designs. The
package reports mechanical sequence checks; downstream biological validation
remains the user's responsibility.
