```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)

docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

# bionemor

Evo 2 is a biological foundation model trained on DNA sequences. ESM-2 is a
biological foundation model trained on protein sequences.

`bionemor` supports Evo 2 and ESM-2 on NVIDIA GPUs from R. Evo 2 can generate
and score DNA, calculate positional profiles and embeddings, and be
[fine-tuned](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
on your own sequences. ESM-2 creates protein embeddings for similarity,
clustering, and downstream models.

Inference inputs are ordinary R character vectors or Biostrings sequence sets.
Generation and scoring return data frames, while pooled embeddings return
numeric matrices. Longer runs can be saved, monitored, and reopened from
another R session.

> Running a model requires a supported CUDA-capable NVIDIA GPU. There is no CPU fallback.
> You can install `bionemor` and inspect the available models without a GPU.

## Install bionemor

```r
pak::pak("t-kalinowski/bionemor")
```

## Quick start: generate DNA with Evo 2

This example takes DNA prompts and predicts continuations with Evo 2. Run it on
a Linux machine with a supported NVIDIA GPU and Docker configured for GPU
access. The [setup guide](https://t-kalinowski.github.io/bionemor/articles/bionemor.html)
covers both local machines and NVIDIA Brev.

First, prepare the package-managed Evo 2 runtime:

```{r runtime}
library(bionemor)

workspace <- "~/workspace/bionemor"
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  workspace = workspace
)
evo2_compute <- bionemo_install(evo2_compute)
```

Then select and prepare the model:

```r
model <- evo2_model("7b", evo2_compute)
```

```{r model-capture, include=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
```

Then generate a continuation for each prompt:

```{r evo2-generate}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]
```

The output above was captured on `r docs_date` with an `r docs_gpu` using
package revision `r substr(docs_revision, 1L, 12L)`.

The first install prepares the runtime, and the first model call downloads and
prepares the checkpoint. Later calls reuse both from the workspace.

The same model can score sequences with `evo2_score()`, calculate positional
profiles with `evo2_profile()`, create embeddings with `evo2_embed()`, and be
fine-tuned with `evo2_finetune()`. See the
[fine-tuning guide](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
for the complete workflow.

## How the pieces fit together

The quick start combines a few package objects that record how operations run.
Saved jobs can be reopened later:

| Concept | Role |
|---|---|
| Model descriptor | Identifies a model family, size, configuration, and optional checkpoint. |
| Recipe | Pins the software environment for a model family. |
| Compute descriptor | Combines a recipe with a workspace and execution settings. |
| Operation | Runs work such as `evo2_generate()` or `esm2_embed()`. |
| Job | Saves an operation's request, logs, state, result, and provenance. |

The quick start uses the defaults `backend = "local"` and
`engine = "container"`. The backend says where commands run: on the current
machine or through Slurm. The engine says how the runtime is supplied: in a
container or in a compatible external environment. The local Docker and Brev
setup guides use these defaults.

## Score and embed DNA with Evo 2

Scoring returns reduced token log probabilities; higher values indicate greater
model likelihood. Pooled embeddings return a numeric matrix whose row names
preserve the input IDs.

```{r evo2-score-embed}
scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores

dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)
dim(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Scores are model likelihoods, not direct measurements of biological function.

## Embed proteins with ESM-2

ESM-2 turns protein sequences into vectors for similarity, clustering, or use
as features in downstream R models:

```{r esm2}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is one last-token, L2-normalized protein embedding. The result behaves
like an ordinary numeric matrix and includes provenance for the model and saved
job.

## Run longer work

Operations that accept `async = TRUE` return a saved job. Save its path to
monitor it or reopen it in a later R session:

```r
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  async = TRUE
)

path <- job_path(run)

# In a new R session:
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

## Learn more

- [Set up a GPU runtime and work on Brev](https://t-kalinowski.github.io/bionemor/articles/bionemor.html)
  covers Docker, model setup, custom images, and the full Evo 2 and ESM-2
  workflows.
- [Fine-tune Evo 2 and retain the checkpoint](https://t-kalinowski.github.io/bionemor/articles/evo2-finetune.html)
  covers preprocessing, LoRA, and fitted inference.
- [Browse the function reference](https://t-kalinowski.github.io/bionemor/reference/index.html)
  for all supported operations and controls.
