---
title: "Evo 2 workflows with BioNeMo Recipes"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Evo 2 workflows with BioNeMo Recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`bionemor` uses the pinned BioNeMo Recipes Evo 2 Megatron runtime. R constructs
typed objects, writes JSONL, FASTA, YAML, and run manifests, and launches the
recipe out of process. MBridge is the native checkpoint format.

NIM is not used. The package does not connect to NIM endpoints, pull NIM
images, or apply the NIM hardware matrix.

## Install or select the runtime

The default container is built from the verified upstream recipe Dockerfile at
the locked commit. Its base image is bound to the release-lock digest before
the package helper and provenance labels are appended.

The local container path requires Git, `tar`, and an NVIDIA GPU exposed through
Docker.
Authenticate Docker to `nvcr.io` with an NGC API key before installing:

```{bash ngc-login, eval = FALSE}
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

`bionemo_install()` pulls the locked NGC PyTorch base image and runs GPU-backed
capability and command probes. It is not a CPU-only setup step.

```{r install, eval = FALSE}
library(bionemor)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = normalizePath("~/evo2-work", mustWork = FALSE)
)

bionemo_install_plan(compute)
compute <- bionemo_install(compute)
bionemo_doctor(compute, target = "all")
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package. `bionemo_install()` verifies that environment; it does not create
a Python environment.

## Prepare a checkpoint

The recommended path converts an open Arc Savanna checkpoint to MBridge:

```{r checkpoint, eval = FALSE}
base <- evo2("7b")

checkpoint <- evo2_checkpoint(
  base,
  source = "recommended",
  format = "savanna",
  path = "checkpoints/evo2-7b-mbridge",
  compute = compute
)

model <- evo2("7b", checkpoint = checkpoint)
```

Constructing `evo2()` is offline. Checkpoint download and conversion happen
only in `evo2_checkpoint()`. MBridge registration requires distributed
checkpoint metadata and at least one non-empty weight shard. Recommended
tokenizers resolve inside the selected recipe runtime.

## Generation and scoring

`evo2_generate()` writes all prompts to one JSONL file and loads the model once
for the operation:

```{r inference, eval = FALSE}
generated <- evo2_generate(
  model,
  c(promoter_a = "ACGTACGT", promoter_b = "GCTAGCTA"),
  compute = compute,
  num_tokens = 64L,
  seed = 1L
)

scores <- evo2_score(
  model,
  c(reference = "ACGTACGT", variant = "ACGTTCGT"),
  compute = compute,
  reduction = "mean",
  strand = "both"
)

embeddings <- evo2_embed(
  model,
  c(reference = "ACGTACGT", variant = "ACGTTCGT"),
  compute = compute,
  pool = "mean"
)
```

Generation and scores return ordinary R data frames. Pooled embeddings return
a numeric matrix. Temporary PyTorch prediction tensors are converted inside
the recipe environment and are not public results.

## Fine-tuning

```{r fine-tuning, eval = FALSE}
data <- evo2_dataset(
  train = "data/train.fa",
  validation = "data/validation.fa",
  test = "data/test.fa"
)

prepared <- evo2_prepare(
  data,
  model,
  compute,
  path = "datasets/splice"
)

run <- evo2_finetune(
  model,
  prepared,
  compute = compute,
  steps = 500L,
  method = evo2_lora(rank = 16L, alpha = 32),
  path = "runs/splice-lora",
  async = TRUE
)

job_status(run)
job_logs(run, tail = 50L)
fitted <- job_wait(run)
```

The LoRA checkpoint manifest records its dense base checkpoint. Inference
fails before launch when that base is unavailable.

The documented original 1B fine-tuning path uses BF16. The pinned
`train_evo2` command has no Vortex-style training mode, so MBridge checkpoints
configured for Vortex-style FP8 and the original 20B and 40B sources are
rejected before launch. FP8 delayed scaling is not offered. Full fine-tuning
cannot start from a LoRA checkpoint.

## Durable runs

Each operation stores its request, structured command plan, state, logs,
inputs, portable outputs, and provenance in one run directory:

```{r reopen, eval = FALSE}
path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_result(same_run)
```

A timeout passed to `job_wait()` only stops waiting. Call `job_cancel()` to
cancel the external process, container, or Slurm allocation.
