---
title: "Evo 2 workflows with BioNeMo Recipes"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Evo 2 workflows with BioNeMo Recipes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`bionemor` prepares Evo 2 models, runs inference, and fine-tunes with NVIDIA
BioNeMo Recipes without leaving R. Inputs are R vectors, data frames, or FASTA
files. Results are R data frames, matrices, artifacts, or durable job objects.
You do not need to write Python or interact with Python objects.

Install the development package with:

```{r package-install, eval = FALSE}
pak::pak("t-kalinowski/bionemor")
```

## Set up the runtime once

The default local setup builds the verified upstream Evo 2 recipe container at
the package's locked revision.

The local container path requires Git, `tar`, and an NVIDIA GPU exposed through
Docker.
Authenticate Docker to `nvcr.io` with an NGC API key before installing:

```{bash ngc-login, eval = FALSE}
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

`bionemo_install()` pulls the locked NGC PyTorch base image and runs GPU-backed
capability and command probes. It is not a CPU-only setup step.

```{r install, eval = FALSE}
library(bionemor)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = normalizePath("~/evo2-work", mustWork = FALSE)
)

plan <- bionemo_install_plan(compute)
plan
as.data.frame(plan)

compute <- bionemo_install(compute)
bionemo_doctor(compute, target = "all")
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package.

## Prepare a model

`evo2_models()` lists the package's pinned model registry and compatibility
metadata without preparing weights:

```{r model-registry, eval = FALSE}
evo2_models()
```

`evo2_model()` prepares the recommended dense MBridge checkpoint and returns a
ready model. A later call reuses a matching complete checkpoint:

```{r checkpoint, eval = FALSE}
model <- evo2_model("7b", compute)
```

Preparation is synchronous and can take substantial time and storage. Use
`evo2()` and `evo2_checkpoint()` when you need a custom source, explicit
conversion controls, or an asynchronous job. In short, `evo2_models()` lists
models, `evo2()` describes one offline, and `evo2_model()` returns one with a
checkpoint attached.

## Run inference

`evo2_generate()` writes all prompts to one JSONL file and loads the model once
for the operation:

```{r inference, eval = FALSE}
generated <- evo2_generate(
  model,
  c(promoter_a = "ACGTACGT", promoter_b = "GCTAGCTA"),
  compute = compute,
  num_tokens = 64L,
  seed = 1L
)

scores <- evo2_score(
  model,
  c(reference = "ACGTACGT", variant = "ACGTTCGT"),
  compute = compute,
  reduction = "mean",
  strand = "both"
)

embeddings <- evo2_embed(
  model,
  c(reference = "ACGTACGT", variant = "ACGTTCGT"),
  compute = compute,
  pool = "mean"
)
```

Generation and scores return ordinary R data frames. Generation includes
optional per-token probabilities and mechanical sequence checks. Scores retain
the forward and reverse reduced log probabilities when both strands are
requested. Pooled embeddings return a numeric matrix whose row names preserve
the input IDs.

## Fine-tune with LoRA

```{r fine-tuning, eval = FALSE}
data <- evo2_dataset(
  train = "data/train.fa",
  validation = "data/validation.fa",
  test = "data/test.fa"
)

prepared <- evo2_prepare(
  data,
  model,
  compute,
  path = "datasets/splice"
)

run <- evo2_finetune(
  model,
  prepared,
  compute = compute,
  steps = 500L,
  method = evo2_lora(rank = 16L, alpha = 32),
  path = "runs/splice-lora",
  async = TRUE
)

job_status(run)
job_logs(run, tail = 50L)
fitted <- job_wait(run)
```

For complete examples, see
[Fine-tune Evo 2 and retain the checkpoint](evo2-finetune.html) and
[Run Evo 2 inference from R](evo2-inference.html).

The LoRA checkpoint manifest records its dense base checkpoint. Inference
fails before launch when that base is unavailable.

The documented original 1B fine-tuning path uses BF16. The pinned
`train_evo2` command has no Vortex-style training mode, so MBridge checkpoints
configured for Vortex-style FP8 and the original 20B and 40B sources are
rejected before launch. FP8 delayed scaling is not offered. Full fine-tuning
cannot start from a LoRA checkpoint.

## Runtime and checkpoint details

BioNeMo Recipes runs in an external process. R writes JSONL, FASTA, YAML,
Parquet, and run manifests, launches the pinned recipe, and converts portable
outputs back to R objects. Python remains inside that recipe runtime rather
than becoming a user-facing API.

MBridge is the native checkpoint format. Registration requires distributed
checkpoint metadata and at least one non-empty weight shard. Recommended
tokenizers resolve inside the selected recipe runtime.

NIM is not used. The package does not connect to NIM endpoints, pull NIM
images, or apply the NIM hardware matrix. Slurm support is experimental.

## Durable runs

Each operation stores its request, structured command plan, state, logs,
inputs, portable outputs, and provenance in one run directory:

```{r reopen, eval = FALSE}
path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
result <- job_wait(same_run)
```

A timeout passed to `job_wait()` only stops waiting. Call `job_cancel()` to
cancel the external process, container, or Slurm allocation.
