---
title: "Evo 2 workflows with BioNeMo Framework"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Evo 2 workflows with BioNeMo Framework}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`bionemor` treats BioNeMo Framework as versioned external compute. The R API
writes validated FASTA and configuration files, invokes a documented CLI, and
materializes results back into R. The package does not expose Python model,
trainer, or tensor objects.

## Supported models and profiles

| Model | Size | Profile | Checkpoints | Fit | Response | Score | Raw |
| --- | --- | --- | --- | ---: | ---: | ---: | ---: |
| Evo 2 | 1B | `bionemo-2.6.3` | Hugging Face, NGC, local | yes | yes | yes | yes |
| Evo 2 | 7B | `bionemo-2.6.3` | Hugging Face, NGC, local | yes | yes | yes | yes |
| Evo 2 | 40B | `bionemo-2.6.3` | Hugging Face, NGC, local | yes | yes | yes | yes |

Representation extraction and parameter-efficient fitting are not implemented
for this profile.

## Local container workflow

This primary workflow uses the public BioNeMo Framework image and a public
Hugging Face checkpoint. It does not require an NGC key.

```{r primary-workflow, eval = FALSE}
library(bionemor)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  image = "nvcr.io/nvidia/clara/bionemo-framework:2.6.3",
  workspace = "/home/ubuntu/bionemor-workspace"
)

bionemo_doctor(compute, target = "training", verbose = FALSE)

spec <- evo2("1b")
base_checkpoint <- evo2_checkpoint(
  spec,
  source = "hf://arcinstitute/savanna_evo2_1b_base",
  path = "checkpoints/evo2-1b-8k",
  compute = compute
)
model <- evo2("1b", checkpoint = base_checkpoint)

baseline <- predict(
  model,
  "ACGTACGTACGT",
  type = "response",
  compute = compute,
  num_tokens = 8L
)

scores <- predict(
  model,
  c(reference = "ACGTACGT", alternate = "ACGTTCGT"),
  type = "score",
  compute = compute,
  reduction = "mean"
)

raw <- predict(
  model,
  "ACGTACGT",
  type = "raw",
  compute = compute
)

control <- evo2_fit_control(
  sequence_length = 8192L,
  learning_rate = 1e-5,
  precision = "bf16"
)
fitted <- generics::fit(
  model,
  data = c(first = "ACGTACGTACGT", second = "TGCATGCATGCA"),
  compute = compute,
  steps = 3L,
  control = control,
  timeout = 300
)

after <- predict(
  fitted,
  "ACGTACGTACGT",
  type = "response",
  compute = compute,
  num_tokens = 8L
)
```

Relative paths resolve under `compute@workspace`. For container execution,
absolute inputs and outputs must also be visible inside that workspace.

## Entitled NGC checkpoints

NGC checkpoints use a separate, explicit path. Set `NGC_CLI_API_KEY`, or
`NGC_API_KEY` for compatibility, before preparing the checkpoint:

```{r ngc-checkpoint, eval = FALSE}
ngc_checkpoint <- evo2_checkpoint(
  evo2("7b"),
  source = "ngc://evo2/7b-1m:1.0",
  path = "checkpoints/evo2-7b-1m-ngc",
  compute = compute
)
```

The NGC resource is fetched with `download_bionemo_data` in the selected
BioNeMo environment. The key is passed only to that child process and is not
stored in commands, provenance, manifests, or logs.

## R, CLI, and Python boundary

The execution contract is:

```text
R API
  -> validated FASTA and YAML files
  -> versioned BioNeMo CLI command
  -> external Python process
  -> checkpoint or prediction artifact
  -> typed R result
```

The `bionemo-2.6.3` profile uses `evo2_convert_to_nemo2`,
`preprocess_evo2`, `train_evo2`, `predict_evo2`, and `infer_evo2`.
Distributed execution, Docker, and Slurm do not run through reticulate.

`type = "response"` returns a named character vector. `type = "score"`
returns a data frame. `type = "raw"` returns a file-backed
`BioNeMoArtifact`, so R never loads a large logits tensor.

## Asynchronous jobs

```{r jobs, eval = FALSE}
job <- predict(
  model,
  sequences,
  type = "score",
  compute = compute,
  async = TRUE
)

job_status(job)
job_logs(job, tail = 100L)
prediction <- job_wait(job)
```

A timeout passed to `job_wait()` only stops waiting. It does not cancel the
external job. Use `job_cancel()` explicitly. The `timeout` passed to `fit()`
instead bounds preprocessing and training as one operation. Finite local
operation timeouts require the GNU `timeout` command.

## Plain checkpoint handoff to nimr

The packages exchange one path and do not import one another. Assume
`fitted_7b` is an Evo 2 7B model returned by `fit()`:

```{r nimr-handoff, eval = FALSE}
checkpoint <- bionemor::checkpoint_path(fitted_7b)

profile <- nimr::nim_profile(
  "evo2",
  variant = "7b",
  service = "self-hosted"
)

deployment <- nimr::nim_deploy(
  profile,
  checkpoint = checkpoint
)
```

The user selects the NIM profile explicitly. `bionemor` does not imply NIM
access, image entitlement, or NGC credentials.
