---
title: "Run Evo 2 inference from R"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Run Evo 2 inference from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This example starts from a fresh R session, prepares or reuses the recommended
dense checkpoint, and runs scoring, generation, and embeddings without leaving
R.

## Recreate the compute description

The runtime and checkpoint must use the same pinned BioNeMo Recipes revision.
`pull = FALSE` avoids another registry pull when the locked image is already
installed, while retaining the image-label and runtime-command checks.

```{r runtime}
library(bionemor)

workspace <- normalizePath("~/evo2-work", mustWork = TRUE)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)
compute <- bionemo_install(compute, pull = FALSE)
```

## Prepare or reuse the model

```{r checkpoint}
model <- evo2_model("7b", compute)

checks <- as.data.frame(
  bionemo_doctor(
    compute,
    model = model,
    target = "inference",
    verbose = FALSE
  )
)
stopifnot(!any(checks$status == "fail"))
checks
```

The first call prepares the registry-recommended dense MBridge checkpoint. A
later call reuses it only when the manifest matches the selected model and
pinned recipe runtime. To attach a custom or fitted checkpoint instead, pass
its root to `evo2("7b", checkpoint = "/path/to/checkpoint")`.

## Score sequences

```{r score}
sequences <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

scores <- evo2_score(
  model,
  sequences,
  compute = compute,
  reduction = "mean",
  strand = "forward"
)

stopifnot(
  nrow(scores) == length(sequences),
  identical(scores$id, names(sequences)),
  all(is.finite(scores$score)),
  all(is.finite(scores$forward_score)),
  all(is.na(scores$reverse_score))
)
scores
```

`score`, `forward_score`, and `reverse_score` are reduced log probabilities.
With both strands, `score` is the arithmetic mean of the forward and reverse
values.

## Generate continuations

```{r generate}
generated <- evo2_generate(
  model,
  sequences,
  compute = compute,
  num_tokens = 8L,
  seed = 1L
)

stopifnot(
  nrow(generated) == length(sequences),
  all(generated$generated_tokens == 8L),
  all(nzchar(generated$completion))
)
generated[c("input_id", "prompt", "completion", "finish_reason")]
```

Generation currently supports one sample per prompt. Use a positive seed and
set at most one of `top_k` and `top_p` to a positive value. The result also
contains token counts, optional `log_probabilities` and `probabilities` list
columns, mechanical sequence summaries, and a `validation_warnings` list
column.

## Extract embeddings

```{r embed}
embeddings <- evo2_embed(
  model,
  sequences,
  compute = compute,
  pool = "mean",
  strand = "both"
)

stopifnot(
  identical(rownames(embeddings), names(sequences)),
  all(startsWith(colnames(embeddings), "dim_")),
  all(is.finite(embeddings))
)
```

Pooling is applied within each strand. With `strand = "both"`, the forward and
reverse pooled vectors are averaged. With `pool = "none"`, `evo2_embed()`
writes a Parquet artifact with `id`, `position`, `embedding`, and `strand`
columns.

## Reopen an asynchronous inference run

For a longer batch, save the run path before leaving the R session:

```{r async}
run <- evo2_generate(
  model,
  sequences,
  compute = compute,
  num_tokens = 8L,
  seed = 1L,
  name = "batch-generation",
  async = TRUE
)

run_path <- job_path(run)
writeLines(
  run_path,
  file.path(workspace, "inference-run.txt")
)

# Run this in a later R session with the same workspace.
run_path <- readLines(
  file.path(workspace, "inference-run.txt"),
  n = 1L,
  warn = FALSE
)
same_run <- bionemo_job(run_path)
job_status(same_run)
job_logs(same_run, tail = 50L)
generated <- job_wait(same_run, timeout = 600)
```

Every run keeps its request, command plan, logs, portable outputs, and
provenance under `.bionemor/runs` in the compute workspace.
