---
title: "Fine-tune Evo 2 and retain the checkpoint"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Fine-tune Evo 2 and retain the checkpoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

This example prepares a small dataset, runs two LoRA optimizer steps, retains
the fitted MBridge checkpoint, and loads it for a generation and scoring smoke
check. It exercises the training and checkpoint-loading path; two steps on
repeated toy sequences do not produce a useful biological model.

The example requires an NVIDIA GPU, the locked BioNeMo Recipes runtime, and
enough storage for the Evo 2 7B dense checkpoint. The opt-in Brev acceptance
workflow requests one NVIDIA L40S and records the exact runtime and hardware
only after the complete workflow succeeds.

## Configure the runtime and dense checkpoint

```{r runtime}
library(bionemor)

workspace <- normalizePath("~/evo2-work", mustWork = FALSE)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)
compute <- bionemo_install(compute)

model <- evo2_model(
  "7b",
  compute,
  path = "checkpoints/evo2-7b-mbridge"
)

training_checks <- as.data.frame(
  bionemo_doctor(
    compute,
    model = model,
    target = "training",
    verbose = FALSE
  )
)
stopifnot(!any(training_checks$status == "fail"))
training_checks
```

`evo2_model()` reuses a complete destination when its manifest matches the
recommended source and locked recipe. Use `evo2()` and `evo2_checkpoint()` for
custom checkpoint sources or asynchronous preparation.

## Prepare a tiny dataset

The nucleotide tokenizer uses one token per base. Each input below contains
127 bases, and preprocessing appends the end-of-document token to produce the
128-token samples requested by the fitting control.

```{r data}
sequence_127 <- function(pattern) {
  stopifnot(
    is.character(pattern),
    length(pattern) == 1L,
    !is.na(pattern),
    nzchar(pattern)
  )
  substr(
    strrep(pattern, ceiling(127 / nchar(pattern))),
    1L,
    127L
  )
}

train <- c(
  train_1 = sequence_127("ACGT"),
  train_2 = sequence_127("TGCA"),
  train_3 = sequence_127("GATTACA"),
  train_4 = sequence_127("CCGTA")
)
validation <- c(validation_1 = sequence_127("AGCT"))
test <- c(test_1 = sequence_127("TGCAT"))

stopifnot(all(nchar(c(train, validation, test)) == 127L))

data <- evo2_dataset(
  train = train,
  validation = validation,
  test = test
)

prepared <- evo2_prepare(
  data,
  model,
  compute,
  path = "datasets/tiny-evo2-128",
  control = evo2_preprocess_control(
    append_eod = TRUE,
    sample_length = 128L,
    workers = 1L,
    seed = 17L
  )
)
```

Use independent biological sequences and held-out splits for real work. The
explicit validation and test inputs keep this small example deterministic.

## Run two LoRA steps

```{r fit}
run <- evo2_finetune(
  model,
  prepared,
  compute = compute,
  steps = 2L,
  method = evo2_lora(
    rank = 4L,
    alpha = 8,
    dropout = 0
  ),
  control = evo2_fit_control(
    sequence_length = 128L,
    global_batch_size = 1L,
    micro_batch_size = 1L,
    learning_rate = 1e-4,
    minimum_learning_rate = 0,
    warmup_steps = 0L,
    decay_steps = 2L,
    constant_steps = 0L,
    eval_interval = 1L,
    eval_iters = 1L,
    log_interval = 1L,
    precision = "bf16",
    keep_checkpoints = 1L,
    workers = 1L,
    seed = 17L,
    dataset_seed = 17L
  ),
  path = "artifacts",
  name = "tiny-evo2-lora",
  async = TRUE,
  timeout = 3600
)

writeLines(
  job_path(run),
  file.path(workspace, "fine-tune-run.txt")
)

job_status(run)
job_logs(run, tail = 50L)
fitted <- job_wait(run, timeout = 3600)
```

`job_wait()` returns the fitted `Evo2Model`. Its timeout only stops the R
process from waiting; it does not cancel training. Use `job_cancel()` when the
external job itself should stop.

## Record the fitted checkpoint

```{r checkpoint}
fitted_checkpoint <- checkpoint_path(fitted)
manifest <- checkpoint_manifest(fitted)

stopifnot(
  dir.exists(fitted_checkpoint),
  identical(manifest$format, "mbridge"),
  identical(manifest$kind, "lora"),
  dir.exists(manifest$base_checkpoint_path)
)

writeLines(
  fitted_checkpoint,
  file.path(workspace, "fitted-checkpoint.txt")
)
```

The checkpoint already exists on disk when `job_wait()` returns. The text file
is only a convenient pointer for a later R session. Keep the checkpoint root
returned by `checkpoint_path()`, which contains `bionemor-checkpoint.json`;
do not select an `iter_*` directory manually.

A LoRA checkpoint records the exact dense base checkpoint and its digest.
Inference requires both the adapter and that base checkpoint to remain at
their recorded paths.

## Smoke-check fitted inference

```{r smoke}
probe <- c(probe = sequence_127("ACGTGCAA"))

scores <- evo2_score(
  fitted,
  probe,
  compute = compute,
  reduction = "mean",
  strand = "forward"
)

generated <- evo2_generate(
  fitted,
  probe,
  compute = compute,
  num_tokens = 8L,
  seed = 1L
)

stopifnot(
  nrow(scores) == 1L,
  is.finite(scores$score[[1L]]),
  nrow(generated) == 1L,
  generated$generated_tokens[[1L]] == 8L,
  nzchar(generated$completion[[1L]])
)

scores
generated[c("prompt", "completion", "finish_reason")]
```

The smoke check verifies that the fitted checkpoint loads and returns
well-formed results. It deliberately does not require two training steps to
improve a score or generate exact bases.

See [Run Evo 2 inference from R](evo2-inference.html) for the recommended
dense-model workflow.
