---
title: "Run BioNeMo jobs with Slurm"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Run BioNeMo jobs with Slurm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

The Slurm backend submits one-node jobs with `sbatch`, reads allocation state
with `sacct`, and cancels with `scancel`. The workspace must be shared between
the R login process and the allocated node.

## Installed Python environment

Use `engine = "python"` when BioNeMo Framework 2.6.3 and its CLI commands are
already installed on compute nodes:

```{r slurm-python, eval = FALSE}
compute <- bionemo_compute(
  backend = "slurm",
  engine = "python",
  workspace = "/shared/projects/evo2",
  gpus = 4L,
  queue = "gpu",
  account = "biology",
  walltime = "01:00:00"
)

bionemo_doctor(compute, target = "training", verbose = FALSE)
```

## Apptainer

Use `engine = "container"` for an Apptainer image visible on compute nodes:

```{r slurm-apptainer, eval = FALSE}
container_compute <- bionemo_compute(
  backend = "slurm",
  engine = "container",
  image = "/shared/images/bionemo-framework-2.6.3.sif",
  workspace = "/shared/projects/evo2",
  gpus = 4L,
  queue = "gpu"
)
```

The generated allocation runs `apptainer exec --nv`. `bionemor` does not
build the SIF or select a cluster-specific module configuration.

## Submit and control a job

```{r slurm-job, eval = FALSE}
model <- evo2("7b", checkpoint = "/shared/models/evo2-7b-1m")

job <- predict(
  model,
  candidate_sequences,
  type = "score",
  compute = compute,
  async = TRUE
)

job_status(job)               # queries sacct
job_logs(job, tail = 100L)
result <- job_wait(job, timeout = 600)
```

A `job_wait()` timeout does not cancel a Slurm allocation. Call
`job_cancel(job)` to issue `scancel`. Unknown scheduler states are reported
with the original state string.

Checkpoint conversion also uses one `sbatch --wait` allocation. The
destination and any local source checkpoint must be inside the shared
workspace.
