---
title: "Get started with bionemor"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Get started with bionemor}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

`bionemor` supports Evo 2 operations on DNA and ESM-2 protein embeddings through
a shared R interface for configuring compute, running and monitoring jobs,
returning results to R, and recording how they were produced.

This article begins with Evo 2 DNA, then uses ESM-2 to create pooled protein
embeddings through the same R interface.

Six concepts recur throughout the package:

- A *model descriptor* is a lightweight S7 object that identifies a family,
  size, configuration, and optional checkpoint. It does not load weights into
  the R process.
- A *checkpoint* contains model weights stored outside R.
- A *recipe* describes the pinned source, dependencies, and commands used to
  build or verify the model family's runtime.
- An *operation* is requested through a family-specific function such as
  `evo2_score()` or `esm2_embed()`.
- A *compute descriptor* records where jobs run, which recipe they use, and
  where checkpoints and job files are stored. A container is one runtime engine
  selected by compute, not another kind of model object.
- A *job* records a run so it can be monitored or reopened from another R
  session.

These concepts apply to both Evo 2 and ESM-2. Each recipe supplies the
corresponding model commands, while the package provides the shared execution
lifecycle.

The output in this article was captured on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`. The executable source
is `vignettes-src/bionemor.Rmd`; the package vignette is pre-rendered so it can
be read without a GPU.

## Start with an NVIDIA GPU

All current model operations require a supported CUDA-capable NVIDIA GPU.
There is no CPU fallback. Recipe and model metadata can be inspected without a
GPU.

For local use, run R on a Linux machine with Git, `tar`, Docker, and the NVIDIA
Container Toolkit. Configure Docker so containers can see the GPU, and
authenticate Docker to `nvcr.io` before installing a runtime.

NVIDIA NGC is the container registry that supplies the runtime's base image.
Create an NGC API key, then authenticate Docker on the GPU machine:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

If you do not have a suitable machine,
[NVIDIA Brev](https://brev.nvidia.com/) is one way to rent one. Review the
hourly prices shown by `brev search` before creating an instance. Download the
package's setup script first:

```bash
curl --fail --location --output bionemor-brev-setup.sh \
  https://raw.githubusercontent.com/t-kalinowski/bionemor/main/tools/brev/setup.sh
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --mode vm \
  --stoppable --gpu-name L40S --min-vram 48
brev exec bionemor-gpu @bionemor-brev-setup.sh
brev shell bionemor-gpu
```

The setup script installs the current R release with rig, installs `bionemor`
with pak, and creates `~/workspace/bionemor`. R runs directly on the Brev VM;
the BioNeMo recipe runtime runs in Docker with `backend = "local"`. Brev
supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.

Data and checkpoints remain user-managed. Copy local files into the persistent
workspace with the CLI, or download them from the VM:

```bash
brev copy ./data/ \
  bionemor-gpu:/home/ubuntu/workspace/bionemor/data/
```

Files below `~/workspace` persist when a stoppable instance is stopped. Deleting
the instance deletes that storage, so retain another copy of important data.
Creating the instance starts a billable resource. When you finish, leave the
remote shell and stop the instance from your local terminal:

```bash
brev stop bionemor-gpu
```

## Supported models and operations

```{r package}
library(bionemor)
```

The function name identifies the model family and the work being requested:

| Model family | Input | Available operations | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Checkpoint preparation and export, generation, scoring, positional profiles, embeddings, training-data preprocessing, and fine-tuning | `evo2_models()`, `evo2_model()`, `evo2_checkpoint()`, `evo2_export()`, `evo2_preprocess()`, `evo2_generate()`, `evo2_score()`, `evo2_profile()`, `evo2_embed()`, `evo2_finetune()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

Use these family-specific functions directly. Both model families return
ordinary R objects and use the same job functions for longer-running work.

## Install an Evo 2 runtime

Create a compute descriptor for the machine and workspace, then call
`bionemo_install()`. The recipe argument is required: it makes the selected
model family and runtime visible in user code. The first installation downloads
and builds the pinned BioNeMo Recipes runtime, so it can take several minutes.

```{r evo2-runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/workspace/bionemor"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package. `backend = "slurm"` can use that external environment or an
existing Apptainer image on a shared cluster. If setup fails,
`bionemo_doctor()` reports runtime and GPU checks.

### Extend the recipe image

Add site-specific system or Python dependencies by building on the package's
verified image, not directly on the raw NVIDIA base image. After installation,
`evo2_compute@image` prints the managed recipe image. Export the same value in a
shell:

```bash
export BIONEMOR_RECIPE_IMAGE="$(
  Rscript --vanilla -e '
    library(bionemor)
    compute <- bionemo_compute(
      recipe = evo2_recipe(),
      workspace = "~/workspace/bionemor"
    )
    cat(compute@image)
  '
)"
```

Then build a Dockerfile such as:

```dockerfile
ARG BIONEMOR_RECIPE_IMAGE
FROM ${BIONEMOR_RECIPE_IMAGE}

# Add site-specific Dockerfile instructions here.
```

```bash
docker build \
  --build-arg BIONEMOR_RECIPE_IMAGE \
  --tag example/bionemor-evo2:site .
```

Select that local tag explicitly and verify the resulting prebuilt recipe image:

```r
custom_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace,
  image = "example/bionemor-evo2:site"
)
custom_compute <- bionemo_install(custom_compute)
```

`bionemo_install()` inspects an explicit image but does not build or pull it.
The R package stays on the VM; it does not need to be installed in this runtime
image.

## Evo 2 example: prepare or reuse a model

The pinned Evo 2 recipe is built on Megatron Bridge and consumes MBridge-format
checkpoints. `evo2_model()` needs a compute descriptor because it downloads the
selected weights when needed, converts them in that runtime, stores them in the
compute workspace, and returns a model descriptor pointing to the checkpoint.
The prepared checkpoint is reused on later calls from the same workspace.

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r evo2-model-capture, echo=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
model
```

`evo2_models()` lists other registered checkpoints and their hardware
compatibility without downloading them.

## Evo 2 example: score DNA

```{r evo2-score}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores
```

The values are reduced token log probabilities. Higher values mean that the
model assigns the sequence greater likelihood; they are not calibrated
measurements of biological function. With `strand = "both"`, each strand is
reduced independently and `score` is their arithmetic mean.

## Evo 2 example: generate DNA

```{r evo2-generate}
generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]
```

Generation continues each DNA prompt. It supports one sample per prompt; set at
most one of `top_k` and `top_p` to a positive value. Generated sequences are
model outputs, not validated biological designs.

## Evo 2 example: extract pooled embeddings

```{r evo2-embed}
dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)

dim(dna_embeddings)
str(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is a numeric representation of one input sequence for downstream
analysis. Pooling excludes padding. With `strand = "both"`, independently
pooled forward and reverse vectors are averaged. Use `pool = "none"` to write
position-level embeddings to a Parquet artifact instead of loading them all
into memory.

## ESM-2 example: embed proteins

`bionemor` also supports pooled protein embeddings with ESM-2. This complete
example configures a compute descriptor, installs the pinned native Transformers
and Transformer Engine runtime, and embeds two short proteins with the smallest
registered model. `esm2_embed()` accepts multiple proteins and returns one
pooled embedding per input. ESM-2 inference currently requires `gpus = 1`; the
runtime processes proteins one at a time on the selected GPU to preserve their
bidirectional attention boundaries.

```{r esm2-embed}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  gpus = 1L,
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

The result contains one last-token, L2-normalized row per protein. It remains a
standard numeric matrix for ordinary R analysis; its `provenance` attribute
records the model, recipe revision, and path where the job was saved. Use these
vectors for sequence similarity, clustering, or as features in downstream R
models. They are model representations, not measurements of protein function.
`esm2_models()` lists the available pinned checkpoints and their embedding
dimensions without downloading weights. Transformers downloads the selected
checkpoint on first use and caches it below the ESM-2 workspace.

## Reopen a job in another R session

Every operation records its request, command plan, logs, portable outputs, and
provenance under `.bionemor/runs` in the compute workspace. For a longer batch,
return a job immediately and save its path:

```{r async, eval=FALSE}
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L,
  async = TRUE
)

path <- job_path(run)

# In a new R session:
same_run <- bionemo_job(path)
job_status(same_run)
job_logs(same_run, tail = 50L)
result <- job_wait(same_run, timeout = 600)
```

A `job_wait()` timeout only stops waiting. Call `job_cancel()` to cancel the
underlying process, container, or Slurm allocation.

To preprocess training data and fine-tune Evo 2, continue with
`vignette("evo2-finetune")`.
