---
title: "Evo 2 workflows from R"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Evo 2 workflows from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
docs_run_id <- Sys.getenv("BIONEMOR_DOCS_RUN_ID")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision),
  nzchar(docs_run_id)
)
```

`bionemor` prepares Evo 2 models and runs inference with NVIDIA BioNeMo
Recipes. Inputs are R vectors, data frames, or FASTA files. Results are R data
frames, matrices, artifacts, or durable job objects.

The output in this article was captured on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`. The executable source
is `vignettes-src/bionemor.Rmd`; the package vignette is pre-rendered so it can
be read without a GPU.

## Set up the runtime once

The local container path requires Git, `tar`, Docker, and an NVIDIA GPU exposed
through Docker. Authenticate Docker to `nvcr.io`, then inspect and run the
installation plan:

```{r runtime}
library(bionemor)

workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/evo2-work"),
  mustWork = FALSE
)
compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)

bionemo_install_plan(compute)
compute <- bionemo_install(compute)
doctor <- bionemo_doctor(compute, target = "inference", verbose = FALSE)
stopifnot(doctor@ok)
doctor
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package.

## Prepare or reuse the model

`evo2_models()` lists the package registry without preparing weights.
`evo2_model()` prepares the recommended dense MBridge checkpoint and binds the
compute descriptor to the returned model:

```{r model}
models <- evo2_models(compute, compatible = TRUE)
models[models$name %in% c("1b-base", "7b"), c(
  "name", "parameters", "context_length", "compatible"
)]

checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = compute
)
model
```

The capture reuses the prepared checkpoint named by
`BIONEMOR_DOCS_CHECKPOINT`. In a new workspace, replace the `evo2()` call with
`model <- evo2_model("7b", compute)` to prepare the recommended source. Both
paths return a compute-bound model, so inference calls do not repeat `compute`.

## Score sequences

```{r score}
sequences <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

scores <- evo2_score(
  model,
  sequences,
  reduction = "mean",
  strand = "both"
)
scores[c("id", "score", "forward_score", "reverse_score")]
```

The values are reduced token log probabilities. Higher values indicate greater
model likelihood. With `strand = "both"`, each strand is reduced independently
and `score` is their arithmetic mean.

## Generate continuations

```{r generate}
generated <- evo2_generate(
  model,
  sequences,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]
```

Generation supports one sample per prompt. Set at most one of `top_k` and
`top_p` to a positive value. The full result also contains token counts,
optional probability list columns, mechanical sequence summaries, and
`validation_warnings`.

## Extract pooled embeddings

```{r embed}
embeddings <- evo2_embed(
  model,
  sequences,
  pool = "mean",
  strand = "both"
)

dim(embeddings)
round(embeddings[, 1:4, drop = FALSE], 4)
```

Pooling excludes padding. With `strand = "both"`, independently pooled forward
and reverse vectors are averaged. Use `pool = "none"` to write position-level
embeddings to a Parquet artifact.

## Reopen a durable job

Every operation records its request, command plan, logs, portable outputs, and
provenance under `.bionemor/runs` in the compute workspace. For a longer batch,
return a job immediately and save its path:

```{r async, eval=FALSE}
run <- evo2_generate(
  model,
  sequences,
  num_tokens = 8L,
  seed = 17L,
  name = paste0(docs_run_id, "-batch-generation"),
  async = TRUE
)

path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
job_logs(same_run, tail = 50L)
result <- job_wait(same_run, timeout = 600)
```

A `job_wait()` timeout only stops waiting. Call `job_cancel()` to cancel the
underlying process, container, or Slurm allocation.
