---
title: "Get started with BioNeMo workflows from R"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Get started with BioNeMo workflows from R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

`bionemor` runs biological foundation-model workflows from R with NVIDIA
BioNeMo Recipes. BioNeMo Recipes supplies the model-specific training and
inference programs. `bionemor` installs a pinned recipe runtime, prepares
inputs, starts jobs, and returns their results to R.

The package currently supports Evo 2 workflows for DNA sequences and ESM-2
pooled embeddings for protein sequences. Evo 2 has the broadest interface, so
most of this article uses it as a detailed example. ESM-2 then shows how the
same API works with another model family.

Five concepts recur throughout the package:

- A *recipe* selects the model-specific BioNeMo runtime.
- A *workflow* is one model-family operation, such as DNA scoring or protein
  embedding.
- A *compute descriptor* records where jobs run, which recipe they use, and
  where checkpoints and job files are stored.
- A *model* describes a registered checkpoint and can retain its compute
  descriptor.
- A *job* records a run so it can be monitored or reopened from another R
  session.

The output in this article was captured on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`. The executable source
is `vignettes-src/bionemor.Rmd`; the package vignette is pre-rendered so it can
be read without a GPU.

## Start with an NVIDIA GPU

All current model operations require a supported CUDA-capable NVIDIA GPU.
There is no CPU fallback. Workflow and model discovery work without a GPU.

For local use, run R on a Linux machine with Git, `tar`, Docker, and the NVIDIA
Container Toolkit. Configure Docker so containers can see the GPU, and
authenticate Docker to `nvcr.io` before installing a runtime.

NVIDIA NGC is the container registry that supplies the runtime's base image.
Create an NGC API key, then authenticate Docker on the GPU machine:

```bash
echo "$NGC_API_KEY" | docker login nvcr.io \
  --username '$oauthtoken' --password-stdin
```

If you do not have a suitable machine,
[NVIDIA Brev](https://brev.nvidia.com/) is one way to rent one. Review the
hourly prices shown by `brev search` before creating an instance:

```bash
brev search --stoppable --gpu-name L40S --min-vram 48 --sort price
brev create bionemor-gpu --stoppable --gpu-name L40S --min-vram 48
brev shell bionemor-gpu

# Run this from your local shell when you finish.
brev stop bionemor-gpu
```

Creating the instance starts a billable resource. Stop it when you finish, and
check Brev's current pricing and storage policy.

Brev supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.
Run the package with the `local` backend inside the instance.

## Discover supported workflows

`bionemo_workflows()` reports the model families and operations implemented by
the installed package. Discovery is offline and does not download weights:

```{r workflows}
library(bionemor)

workflows <- bionemo_workflows()
workflows[c("id", "family", "task")]
```

| Model family | Input | Supported workflows | Main functions |
|---|---|---|---|
| Evo 2 | DNA sequences | Checkpoint preparation and export, generation, scoring, positional profiles, embeddings, data preparation, and fine-tuning | `evo2_models()`, `evo2_model()`, `evo2_generate()`, `evo2_score()`, `evo2_embed()`, `evo2_finetune()` |
| ESM-2 | Protein sequences | Pooled protein embeddings | `esm2_models()`, `esm2_model()`, `esm2_embed()` |

The family-specific functions are the ordinary R interface. For automation,
`bionemo_run()` also accepts a family-qualified ID such as `"evo2/score"` or
`"esm2/embed"` from this table.

## Install an Evo 2 runtime

Create a compute descriptor for the machine and workspace, then call
`bionemo_install()`. The recipe argument is required: it makes the selected
model family and runtime visible in user code. The first installation downloads
and builds the pinned BioNeMo Recipes runtime, so it can take several minutes.

```{r evo2-runtime}
workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/bionemor-work"),
  mustWork = FALSE
)
evo2_compute <- bionemo_compute(
  recipe = evo2_recipe(),
  backend = "local",
  engine = "container",
  workspace = workspace
)

evo2_compute <- bionemo_install(evo2_compute)
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package. If setup fails, `bionemo_doctor()` reports runtime and GPU checks.

## Evo 2 example: prepare or reuse a model

`evo2_model()` downloads the selected weights when needed, converts them to the
checkpoint format used by the Evo 2 recipe, and binds the compute descriptor to
the returned model. The prepared checkpoint is reused on later calls from the
same workspace.

```r
model <- evo2_model("7b", evo2_compute)
model
```

```{r evo2-model-capture, echo=FALSE}
checkpoint <- Sys.getenv("BIONEMOR_DOCS_CHECKPOINT")
stopifnot(dir.exists(checkpoint))
model <- evo2(
  "7b",
  checkpoint = checkpoint,
  compute = evo2_compute
)
model
```

`evo2_models()` lists other registered checkpoints and their hardware
compatibility without downloading them.

## Evo 2 example: score DNA

```{r evo2-score}
dna <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

scores <- evo2_score(
  model,
  dna,
  reduction = "mean",
  strand = "both"
)
scores
```

The values are reduced token log probabilities. Higher values mean that the
model assigns the sequence greater likelihood; they are not calibrated
measurements of biological function. With `strand = "both"`, each strand is
reduced independently and `score` is their arithmetic mean.

## Evo 2 example: generate DNA

```{r evo2-generate}
generated <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]
```

Generation continues each DNA prompt. It supports one sample per prompt; set at
most one of `top_k` and `top_p` to a positive value. Generated sequences are
model outputs, not validated biological designs.

## Evo 2 example: extract pooled embeddings

```{r evo2-embed}
dna_embeddings <- evo2_embed(
  model,
  dna,
  pool = "mean",
  strand = "both"
)

dim(dna_embeddings)
str(dna_embeddings)
round(dna_embeddings[, 1:4, drop = FALSE], 4)
```

Each row is a numeric representation of one input sequence for downstream
analysis. Pooling excludes padding. With `strand = "both"`, independently
pooled forward and reverse vectors are averaged. Use `pool = "none"` to write
position-level embeddings to a Parquet artifact instead of loading them all
into memory.

## ESM-2 example: embed proteins

ESM-2 uses a separate recipe because its runtime and model commands differ from
Evo 2. This complete example creates a second compute descriptor, installs the
pinned native Transformers and Transformer Engine runtime, and embeds two short
proteins with the smallest registered model.

Installation uses the recipe's native Transformers and Transformer Engine path
and does not compile vLLM. ESM-2 inference currently requires `gpus = 1`. You
may supply multiple proteins; the runtime processes them one at a time on the
selected GPU to preserve their bidirectional attention boundaries.

```{r esm2-embed}
esm2_compute <- bionemo_compute(
  recipe = esm2_recipe(),
  backend = "local",
  engine = "container",
  gpus = 1L,
  workspace = file.path(workspace, "esm2")
)
esm2_compute <- bionemo_install(esm2_compute)
protein_model <- esm2_model("8m", esm2_compute)

proteins <- c(
  protein_1 = "MKTAYIAKQRQISFVKSHFSRQ",
  protein_2 = "GAVLILKKKGHHEAELKPLAQSHATK"
)
protein_embeddings <- esm2_embed(protein_model, proteins)

dim(protein_embeddings)
str(protein_embeddings)
round(protein_embeddings[, 1:4, drop = FALSE], 4)
```

The result contains one last-token, L2-normalized row per protein. Use these
vectors for sequence similarity, clustering, or as features in downstream R
models. They are model representations, not measurements of protein function.
`esm2_models()` lists the available pinned checkpoints and their embedding
dimensions without downloading weights. Transformers downloads the selected
checkpoint on first use and caches it below the ESM-2 workspace.

## Reopen a durable job

Every operation records its request, command plan, logs, portable outputs, and
provenance under `.bionemor/runs` in the compute workspace. For a longer batch,
return a job immediately and save its path:

```{r async, eval=FALSE}
run <- evo2_generate(
  model,
  dna,
  num_tokens = 8L,
  seed = 17L,
  async = TRUE
)

path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
job_logs(same_run, tail = 50L)
result <- job_wait(same_run, timeout = 600)
```

A `job_wait()` timeout only stops waiting. Call `job_cancel()` to cancel the
underlying process, container, or Slurm allocation.

To prepare data and train Evo 2 adapters, continue with
`vignette("evo2-finetune")`. For a shared GPU cluster, see
`vignette("slurm")`.
