---
title: "Get started with BioNeMo and Evo 2"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Get started with BioNeMo and Evo 2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(
  nzchar(docs_gpu),
  nzchar(docs_date),
  nzchar(docs_revision)
)
```

`bionemor` lets you run NVIDIA BioNeMo workflows from R. BioNeMo Recipes is
the Python runtime that supplies the model-specific training and inference
commands; `bionemor` installs that runtime, prepares inputs, starts jobs, and
returns their results to R.

The first supported model family is Evo 2, a model for DNA sequences. This
article shows how to score sequences, generate continuations, and extract
embeddings. Three terms recur throughout the package:

- A *checkpoint* is a directory containing learned model weights and metadata.
- A *compute descriptor* tells `bionemor` where to run, which runtime to use,
  and where to keep checkpoints and job files.
- An *Evo 2 model* joins a checkpoint to a compute descriptor. Once prepared,
  it can be passed directly to scoring, generation, embedding, and fine-tuning
  functions.

The output in this article was captured on `r docs_date` with an `r docs_gpu`
using package revision `r substr(docs_revision, 1L, 12L)`. The executable source
is `vignettes-src/bionemor.Rmd`; the package vignette is pre-rendered so it can
be read without a GPU.

## Start with an NVIDIA GPU

All current `bionemor` workflows require a CUDA-capable NVIDIA GPU.
There is no CPU fallback.

For local use, run R on a Linux machine with Git, `tar`, Docker, and the NVIDIA
Container Toolkit. Configure Docker so containers can see the GPU, and
authenticate Docker to `nvcr.io` before installing the runtime.

If you do not have a suitable machine, Brev is one way to rent one. Review the
hourly prices shown by `brev search` before creating an instance:

```bash
brev search --stoppable --min-total-vram 40 --sort price
brev create bionemor-gpu --stoppable --min-total-vram 40
brev shell bionemor-gpu

# Run this from your local shell when you finish.
brev stop bionemor-gpu
```

Creating the instance starts a billable resource. Stop it when you finish, and
check Brev's current pricing and storage policy.

Brev supplies the Linux GPU machine; it is not a `bionemo_compute()` backend.
Run the package with the `local` backend inside the instance.

## Install the runtime

Create a compute descriptor for the machine and workspace, then call
`bionemo_install()`. The first installation downloads and builds the pinned
BioNeMo Recipes runtime, so it can take several minutes:

```{r runtime}
library(bionemor)

workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/evo2-work"),
  mustWork = FALSE
)
compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)

compute <- bionemo_install(compute)
```

Use `engine = "external"` for a compatible recipe environment managed outside
the package. If setup fails, `bionemo_doctor()` reports detailed runtime and
GPU checks.

## Prepare or reuse the model

`evo2_model()` downloads the selected weights when needed, converts them to the
checkpoint format used by BioNeMo Recipes, and binds the compute descriptor to
the returned model. Checkpoints are large, but the prepared copy is reused on
later calls from the same workspace.

```{r model}
model <- evo2_model("7b", compute)
model
```

`evo2_models()` lists other registered checkpoints and their hardware
compatibility without downloading them.

## Score sequences

```{r score}
sequences <- c(
  reference = "ACGTACGTACGT",
  variant = "ACGTACGTTCGT"
)

scores <- evo2_score(
  model,
  sequences,
  reduction = "mean",
  strand = "both"
)
scores
```

The values are reduced token log probabilities. Higher values mean that the
model assigns the sequence greater likelihood; they are not calibrated
measurements of biological function. With `strand = "both"`, each strand is
reduced independently and `score` is their arithmetic mean.

## Generate continuations

```{r generate}
generated <- evo2_generate(
  model,
  sequences,
  num_tokens = 8L,
  seed = 17L
)
generated[c(
  "input_id", "prompt", "completion", "generated_tokens", "finish_reason"
)]
```

Generation continues each DNA prompt. It supports one sample per prompt; set at
most one of `top_k` and `top_p` to a positive value. Generated sequences are
model outputs, not validated biological designs.

## Extract pooled embeddings

```{r embed}
embeddings <- evo2_embed(
  model,
  sequences,
  pool = "mean",
  strand = "both"
)

dim(embeddings)
str(embeddings)
round(embeddings[, 1:4, drop = FALSE], 4)
```

Each row is a numeric representation of one input sequence for downstream
analysis. Pooling excludes padding. With `strand = "both"`, independently
pooled forward and reverse vectors are averaged. Use `pool = "none"` to write
position-level embeddings to a Parquet artifact instead of loading them all
into memory.

## Reopen a durable job

Every operation records its request, command plan, logs, portable outputs, and
provenance under `.bionemor/runs` in the compute workspace. For a longer batch,
return a job immediately and save its path:

```{r async, eval=FALSE}
run <- evo2_generate(
  model,
  sequences,
  num_tokens = 8L,
  seed = 17L,
  async = TRUE
)

path <- job_path(run)
same_run <- bionemo_job(path)
job_status(same_run)
job_logs(same_run, tail = 50L)
result <- job_wait(same_run, timeout = 600)
```

A `job_wait()` timeout only stops waiting. Call `job_cancel()` to cancel the
underlying process, container, or Slurm allocation.

To train adapters on your own sequences, continue with
`vignette("evo2-finetune")`. For a shared GPU cluster, see
`vignette("slurm")`.
