---
title: "Fine-tune Evo 2 with LoRA"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Fine-tune Evo 2 with LoRA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  error = FALSE,
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(nzchar(docs_gpu), nzchar(docs_date), nzchar(docs_revision))
```

This article adapts Evo 2 to a DNA sequence collection with low-rank adaptation
(LoRA), saves the fitted checkpoint, and uses it for scoring and generation.
LoRA trains small adapter matrices while leaving most of the base model
unchanged, so it requires less GPU memory than updating every model weight.

Complete the setup in `vignette("bionemor")` first. Fine-tuning still requires
an NVIDIA CUDA GPU, an installed BioNeMo Recipes runtime, and a prepared Evo 2
checkpoint. The example below uses repeated toy sequences and only two
optimizer steps to exercise the workflow. It does not produce a useful
biological model.

The output was captured on `r docs_date` with an `r docs_gpu` using package
revision `r substr(docs_revision, 1L, 12L)`. The executable source is
`vignettes-src/evo2-finetune.Rmd`; the package vignette is pre-rendered so it
can be read without a GPU.

## Set up the runtime and base model

```{r runtime}
library(bionemor)

workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/evo2-work"),
  mustWork = FALSE
)
run_id <- Sys.getenv(
  "BIONEMOR_DOCS_RUN_ID",
  format(Sys.time(), "docs-%Y%m%d-%H%M%S")
)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)
compute <- bionemo_install(compute)
model <- evo2_model("7b", compute)
```

`evo2_model()` reuses the checkpoint when it is already present in the
workspace.

## Prepare a tiny dataset

`evo2_dataset()` accepts named sequence vectors or FASTA files. Supply
independent validation and test sequences so training and evaluation do not
measure memorization of closely related inputs.

The nucleotide tokenizer uses one token per base. Each toy sequence below has
127 bases; preprocessing appends an end-of-document token to create the
128-token samples requested by the fitting control.

```{r data}
sequence_127 <- function(pattern) {
  stopifnot(is.character(pattern), length(pattern) == 1L, nzchar(pattern))
  substr(strrep(pattern, ceiling(127 / nchar(pattern))), 1L, 127L)
}

train <- c(
  train_1 = sequence_127("ACGT"),
  train_2 = sequence_127("TGCA"),
  train_3 = sequence_127("GATTACA"),
  train_4 = sequence_127("CCGTA")
)
validation <- c(validation_1 = sequence_127("AGCT"))
test <- c(test_1 = sequence_127("TGCAT"))

data <- evo2_dataset(
  train = train,
  validation = validation,
  test = test
)

prepared <- evo2_prepare(
  data,
  model,
  path = file.path("datasets", paste0(run_id, "-tiny-evo2-128")),
  control = evo2_preprocess_control(
    append_eod = TRUE,
    sample_length = 128L,
    workers = 1L,
    seed = 17L
  ),
  overwrite = TRUE
)

data.frame(
  split = names(prepared@manifest$inputs),
  records = vapply(
    prepared@manifest$inputs,
    function(input) input$records,
    integer(1L)
  ),
  row.names = NULL
)
```

Explicit validation and test inputs keep this small example deterministic. For
real work, choose held-out splits that reflect the biological question and
avoid leakage between related sequences.

## Choose LoRA targets

`evo2_lora()` describes which parts of the model receive trainable adapters:

- `"hyena"` selects the Hyena sequence-mixing blocks.
- `"attention"` selects attention blocks.
- `"mlp"` selects feed-forward blocks.

These are adapter locations, not response variables. `rank` controls adapter
capacity, and the effective adapter scale is `alpha / rank`.

```{r method}
method <- evo2_lora(
  rank = 4L,
  alpha = 8,
  dropout = 0,
  targets = c("hyena", "attention", "mlp")
)
method
```

## Fine-tune the model

```{r fit}
run <- evo2_finetune(
  model,
  prepared,
  steps = 2L,
  method = method,
  control = evo2_fit_control(
    sequence_length = 128L,
    global_batch_size = 1L,
    micro_batch_size = 1L,
    learning_rate = 1e-4,
    minimum_learning_rate = 0,
    warmup_steps = 0L,
    decay_steps = 2L,
    constant_steps = 0L,
    eval_interval = 1L,
    eval_iters = 1L,
    log_interval = 1L,
    precision = "bf16",
    keep_checkpoints = 1L,
    workers = 1L,
    seed = 17L,
    dataset_seed = 17L
  ),
  path = file.path("artifacts", run_id, "vignette-lora"),
  name = paste0(run_id, "-vignette-lora"),
  async = TRUE,
  timeout = 3600
)
run

fitted <- job_wait(run, timeout = 3600)
job_status(run, refresh = TRUE)
fitted
```

With `async = TRUE`, `evo2_finetune()` returns immediately with a durable job.
`job_wait()` returns a fitted `Evo2Model` that points to the saved checkpoint
and retains the compute descriptor. A wait timeout does not cancel training.

## Inspect and reuse the fitted checkpoint

```{r checkpoint}
fitted_checkpoint <- checkpoint_path(fitted)
manifest <- checkpoint_manifest(fitted)

data.frame(
  format = manifest$format,
  method = manifest$kind,
  base_checkpoint = basename(manifest$base_checkpoint_path)
)
```

Keep the checkpoint root returned by `checkpoint_path()`, which contains
`bionemor-checkpoint.json`. A LoRA checkpoint records its dense base checkpoint
and requires that base to remain available. Pass `fitted` directly to inference
functions while working in the same R session.

```{r fitted-inference}
probe <- c(probe = sequence_127("ACGTGCAA"))

fitted_scores <- evo2_score(
  fitted,
  probe,
  reduction = "mean",
  strand = "forward"
)
fitted_scores[c("id", "score", "forward_score")]

fitted_generation <- evo2_generate(
  fitted,
  probe,
  num_tokens = 8L,
  seed = 17L
)
fitted_generation[c("input_id", "prompt", "completion", "finish_reason")]
```

These calls confirm that the adapter and dense base load together and return
well-formed results. The values only validate the workflow; two toy training
steps are not evidence of improved biological performance.
