---
title: "Fine-tune Evo 2 and retain the checkpoint"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Fine-tune Evo 2 and retain the checkpoint}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
options(width = 92)
docs_gpu <- Sys.getenv("BIONEMOR_DOCS_GPU")
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(nzchar(docs_gpu), nzchar(docs_date), nzchar(docs_revision))
```

This example prepares a small dataset, runs two LoRA optimizer steps, retains
the fitted MBridge checkpoint, and uses it for generation and scoring. It
exercises the training and checkpoint-loading path; two steps on repeated toy
sequences do not produce a useful biological model.

The output was captured on `r docs_date` with an `r docs_gpu` using package
revision `r substr(docs_revision, 1L, 12L)`. The executable source is
`vignettes-src/evo2-finetune.Rmd`; the package vignette is pre-rendered so it
can be read without a GPU.

## Configure the runtime and dense checkpoint

```{r runtime}
library(bionemor)

workspace <- normalizePath(
  Sys.getenv("BIONEMOR_DOCS_WORKSPACE", "~/evo2-work"),
  mustWork = FALSE
)
run_id <- Sys.getenv(
  "BIONEMOR_DOCS_RUN_ID",
  format(Sys.time(), "docs-%Y%m%d-%H%M%S")
)

compute <- bionemo_compute(
  backend = "local",
  engine = "container",
  workspace = workspace
)
compute <- bionemo_install(compute)
model <- evo2_model("7b", compute)

bionemo_doctor(
  compute,
  model = model,
  target = "training",
  verbose = FALSE
)
```

`evo2_model()` reuses a complete destination only when its manifest matches the
recommended source and pinned recipe. The returned model retains `compute`.

## Prepare a tiny dataset

The nucleotide tokenizer uses one token per base. Each input below contains 127
bases, and preprocessing appends the end-of-document token to produce the
128-token samples requested by the fitting control.

```{r data}
sequence_127 <- function(pattern) {
  stopifnot(is.character(pattern), length(pattern) == 1L, nzchar(pattern))
  substr(strrep(pattern, ceiling(127 / nchar(pattern))), 1L, 127L)
}

train <- c(
  train_1 = sequence_127("ACGT"),
  train_2 = sequence_127("TGCA"),
  train_3 = sequence_127("GATTACA"),
  train_4 = sequence_127("CCGTA")
)
validation <- c(validation_1 = sequence_127("AGCT"))
test <- c(test_1 = sequence_127("TGCAT"))

data <- evo2_dataset(
  train = train,
  validation = validation,
  test = test
)

prepared <- evo2_prepare(
  data,
  model,
  path = file.path("datasets", run_id, "tiny-evo2-128"),
  control = evo2_preprocess_control(
    append_eod = TRUE,
    sample_length = 128L,
    workers = 1L,
    seed = 17L
  ),
  overwrite = TRUE
)
prepared
```

Explicit validation and test inputs keep this small example deterministic. Use
independent biological sequences and held-out splits for real work.

## Choose LoRA targets

`targets` selects semantic module groups in every matching model layer:

- `"hyena"` selects `dense_projection` and `dense`.
- `"attention"` selects `linear_qkv` and `linear_proj`.
- `"mlp"` selects `linear_fc1` and `linear_fc2`.

These are adapter locations, not prediction targets or response variables. The
effective adapter scale is `alpha / rank`. Modules listed in `fully_trainable`
remain unfrozen without receiving adapters; tied `word_embeddings` and
`output_layer` must be listed together.

```{r method}
method <- evo2_lora(
  rank = 4L,
  alpha = 8,
  dropout = 0,
  targets = c("hyena", "attention", "mlp")
)
method
```

## Run two optimizer steps

```{r fit}
run <- evo2_finetune(
  model,
  prepared,
  steps = 2L,
  method = method,
  control = evo2_fit_control(
    sequence_length = 128L,
    global_batch_size = 1L,
    micro_batch_size = 1L,
    learning_rate = 1e-4,
    minimum_learning_rate = 0,
    warmup_steps = 0L,
    decay_steps = 2L,
    constant_steps = 0L,
    eval_interval = 1L,
    eval_iters = 1L,
    log_interval = 1L,
    precision = "bf16",
    keep_checkpoints = 1L,
    workers = 1L,
    seed = 17L,
    dataset_seed = 17L
  ),
  path = file.path("artifacts", run_id, "vignette-lora"),
  name = paste0(run_id, "-vignette-lora"),
  async = TRUE,
  timeout = 3600
)
run

fitted <- job_wait(run, timeout = 3600)
job_status(run, refresh = TRUE)
fitted
```

`job_wait()` returns a fitted `Evo2Model`. Its timeout only stops waiting; it
does not cancel training.

## Inspect and reuse the fitted checkpoint

```{r checkpoint}
fitted_checkpoint <- checkpoint_path(fitted)
manifest <- checkpoint_manifest(fitted)

data.frame(
  path = fitted_checkpoint,
  format = manifest$format,
  kind = manifest$kind,
  base_checkpoint = manifest$base_checkpoint_path,
  recipe_revision = manifest$recipe_revision
)
```

Keep the checkpoint root returned by `checkpoint_path()`, which contains
`bionemor-checkpoint.json`. A LoRA checkpoint records its dense base checkpoint
and requires that base to remain available.

```{r fitted-inference}
probe <- c(probe = sequence_127("ACGTGCAA"))

scores <- evo2_score(
  fitted,
  probe,
  reduction = "mean",
  strand = "forward"
)
scores[c("id", "score", "forward_score")]

generated <- evo2_generate(
  fitted,
  probe,
  num_tokens = 8L,
  seed = 17L
)
generated[c("input_id", "prompt", "completion", "finish_reason")]
```

The smoke check verifies that the adapter and dense base load and return
well-formed results. It does not require two training steps to improve a score
or generate exact bases.
