---
title: "Run BioNeMo Recipes jobs with Slurm"
output: rmarkdown::html_vignette
vignette: |
  %\VignetteIndexEntry{Run BioNeMo Recipes jobs with Slurm}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", error = FALSE)
docs_date <- Sys.getenv("BIONEMOR_DOCS_RENDER_DATE")
docs_revision <- Sys.getenv("BIONEMOR_DOCS_PACKAGE_REVISION")
stopifnot(nzchar(docs_date), nzchar(docs_revision))
```

The Slurm backend submits one-node jobs with `sbatch`, reads allocation state
with `sacct`, and cancels with `scancel`. The R process, compute nodes, and
container must see the same workspace paths.

This reference was pre-rendered on `r docs_date` from package revision
`r substr(docs_revision, 1L, 12L)`. Its commands were not executed on Brev:
scheduler, shared-filesystem, container-runtime, and allocation behavior must be
validated on the target cluster.

Slurm support is experimental. `bionemo_install()` verifies the runtime from a
synchronous Slurm allocation. It does not run GPU or Apptainer probes on the
login host.

The compute-node environment must provide `bash`, `awk`, `mkfifo`, and the same
`Rscript` path used by the submitting R process. It must also provide
`sha256sum` when a local SIF is used. That R installation must have `jsonlite`
available.

## Administrator-managed external runtime

Use `engine = "external"` when the pinned BioNeMo Recipes commands and package
helper are already available on compute nodes:

```{r slurm-external, eval=FALSE}
library(bionemor)

compute <- bionemo_compute(
  backend = "slurm",
  engine = "external",
  workspace = "/shared/projects/evo2",
  gpus = 4L,
  queue = "gpu",
  account = "biology",
  walltime = "01:00:00"
)

compute <- bionemo_install(compute)
bionemo_doctor(compute, target = "training", verbose = FALSE)
```

## Apptainer

Use `engine = "container"` with an existing Apptainer image built from the
verified Evo 2 recipe:

```{r slurm-apptainer, eval=FALSE}
compute <- bionemo_compute(
  backend = "slurm",
  engine = "container",
  image = "/shared/images/bionemor-evo2-e8e7f597.sif",
  workspace = "/shared/projects/evo2",
  gpus = 4L,
  queue = "gpu"
)

compute <- bionemo_install(compute)
```

For a local SIF path, `bionemo_install()` computes and records its SHA-256
digest before submitting verification probes. The image must be readable on the
shared filesystem. A digest-qualified image URI records the supplied digest.

Version 1 does not build a SIF on a compute node or select cluster modules. The
allocation environment must provide `apptainer`.

## Submit and reopen a job

```{r slurm-job, eval=FALSE}
model <- evo2("7b", checkpoint = "/shared/models/evo2-7b-mbridge")

job <- evo2_score(
  model,
  candidate_sequences,
  compute = compute,
  async = TRUE
)

path <- job_path(job)
job_status(job)
job_logs(job, tail = 100L)

job <- bionemo_job(path)
result <- job_wait(job, timeout = 600)
```

A `job_wait()` timeout does not cancel the allocation. Call `job_cancel()` to
issue `scancel`.
